Towards Practical First-Order Model Counting

### 2024-04-18 doc/paper.tex
First draft of the beginning of Section 3

\crefname{line}{line}{lines}
\crefalias{formula}{equation}
\crefname{formula}{Clause}{Clauses}
\creflabelformat{formula}{#2\textup{(#1)}#3}

@@

\DeclareMathOperator{\Dom}{Dom}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator*{\argmin}{argmin}

@@

\todo[inline]{Reorder the Logic subsection to cover \FO{} before comparing it to
  other logics (and clarify the differences between the input format and the
  internal format)}

\paragraph{Three Types of Logics}
\begin{itemize}
  \item See \cref{tbl:logics} for a detailed comparison. The notation introduced
        in the table is standard for \Ctwo{}, new for \UFO{}, and redefined to
        be more specific for \FO{}.
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item In many-sorted logic, each term is assigned to a \emph{sort}, and each
        predicate $p/n$ is assigned to a sequence of $n$ sorts. Each sort has
        its corresponding domain. In the input formula, all domains are assumed
        to be pairwise disjoin.t Most of these assignments are typically left
        implicit and can be reconstructed from the quantifiers. For instance,
        $\forall x,y \in \Delta\text{. }P(x, y)$ implies that variables $x$ and
        $y$ have the same sort. On the other hand, $\forall x \in \Delta\text{.
        }\forall y \in \Gamma\text{. } P(x, y)$ implies that $x$ and $y$ have
        different sorts, and it would be improper to have $x = y$ as part of a
        formula.
  \item
        \begin{itemize}
          \item \FO{} is used as the input format for
                \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
                and its extensions
                \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}~\cite{DBLP:conf/kr/DilkasB23}
                and \Cranetwo{}.
          \item \Ctwo{} is discussed in the literature on
                \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} and related
                methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}
          \item \UFO{} is the input format supported by a version of
                \textsc{FastWFOMC} obtained directly from the authors. Note that
                the publicly available
                version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
                does not support any cardinality constraints.
        \end{itemize}
  \item Note that, in the case of \textsc{ForcLift} and its extensions, support
        for a formula as valid input does not imply that the algorithm will be
        able to compile the formula into a circuit or graph suitable for lifted
        model counting. However, it is known that \textsc{ForcLift} compilation
        is guaranteed to succeed on any \FO{} formula without constants and with
        at most two variables~\cite{DBLP:conf/nips/Broeck11}.
\end{itemize}

\paragraph{The \FO{} Logic (After Skolemization)}
\begin{itemize}
  \item Although existential quantifiers are supported, here we describe the
        format used internally. During preprocessing, all existential
        quantifiers are eliminated using
        Skolemization~\cite{DBLP:conf/kr/BroeckMD14}, and the formula is
        rewritten into a conjunction of clauses, each of which is in
        \emph{prenex normal form}~\cite{hinman2018fundamentals}.
  \item In the spirit of keeping sorts implicit, we always assume formulas `type
        check' with respect to sorts. For example, if $P(x)$, $P(y)$, and
        $x \ne c$ are all part of the formula (for some predicate $P$, variables
        $x$ and $y$, and constant $c$), then $x$, $y$, and $c$ all have the same
        sort.
  \item A \emph{formula} is a conjunction of clauses.
  \item A \emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
        }\forall x_{2} \in \Delta_{2}\dots\text{
        }\forall x_{n} \in \Delta_{n}\text{. }\phi(x_{1}, x_{2}, \dots, x_{n})$,
        where $\phi$ is a disjunction of literals that only contain variables
        $x_{1}, \dots, x_{n}$ (and any constants).
  \item We say that a clause is a \emph{(positive) unit clause} if:
        \begin{itemize}
          \item there is only one literal with a predicate, and
          \item it is a positive literal.
        \end{itemize}
  \item A \emph{literal} is either an atom (i.e., a \emph{positive} literal) or
        its negation (i.e., a \emph{negative} literal).
  \item An \emph{atom} is either:
  \begin{itemize}
    \item $P(t_{1}, \dots, t_{m})$ for some predicate $P/m$ and terms
          $t_{1}, \dots, t_{m}$ or
    \item $x=y$ for some terms $x$ and $y$
  \end{itemize}
  \item An atom is \emph{ground} if it contains no variables (i.e., only
        constants).
  \item The \emph{arity} of a predicate is the number of arguments it takes,
        i.e., $m$ in the case of predicate $P$.
  \item When we want to denote a predicate together with its arity, we write
        $P/m$.
  \item A \emph{term} is either a variable or a constant.
  \item Throughout the paper, we use set-theoretic notation, interpreting a
        formula as a set of clauses and a clause as a set of literals. Moreover,
        for readability, clauses written on separate lines are implicitly
        conjoined.
\end{itemize}

For any variable $v$, let $\Dom(v)$ denote the domain over which $v$ is
quantified. For any clause or literal $l$, let $\Vars(l)$ denote the set of
variables in $l$, and, similarly, let
$\Doms(l) \coloneqq \{\, \Dom(v) \mid v \in \Vars(l) \,\}$.

@@

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item the definition of a function is a strong assumption
    \item the notation for a function call is now outdated
  \end{itemize}
}

\begin{itemize}
  \item We write $\expr{}$ for an arbitrary algebraic expression.
  \item A few terms have different meanings in the context of algebra as
        compared to logic:
        \begin{itemize}
          \item a \emph{variable} is either:
                \begin{itemize}
                  \item a parameter of a function or
                  \item a variable introduced by a summation, e.g., $i$ in
                        $\sum_{i=1}^{n} \expr$.
                \end{itemize}
          \item A \emph{constant} is a non-negative integer.
        \end{itemize}
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{\protect\CompileWithBaseCases{$\phi$}}\label{alg:main}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;
  $\mathcal{E} \gets \Simplify{$\mathcal{E}$}$\;
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i} \in \mathbb{N}_{0}$}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $\mathcal{E} \gets \mathcal{E} \cup \CompileWithBaseCases{$\psi$}$\;
  }
\end{algorithm}

\todo[inline]{In \cref{alg:main}, some kind of translation also needs to take
  place, right?}

We write \Crane for the first-order knowledge compilation algorithm of the
original \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} that compiles a formula
$\phi$ into:
\begin{itemize}
  \item a set $\mathcal{E}$ of equations,
  \item a map $\mathcal{F}$ from function names to formulas, and
  \item a map $\mathcal{D}$ from function names and parameter indices to
        domains.
\end{itemize}

\todo[inline,caption={}]{
  \begin{itemize}
    \item \textsc{Crane}, \Cranetwo{}, or \Crane in the paragraph below
  \end{itemize}
}

See \cref{alg:main}. The evaluation of base cases is done by simplifying the
clauses and then using \textsc{Crane} to find the base cases. A particular
domain is selected, and the clauses are simplified. Then, \textsc{Crane} is
called on those clauses to evaluate the base cases. After that, we change the
function names and variable to make it consistent with the previous domain to
variable mapping, and append these base cases to the set of equations.

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item later check if I'm still using the syntax for $\Vars$, $\Dom$, etc.
    \item align the last line to be a bit neater
    \item \Cref{alg2} now supports any base case domain sizes
    \item NOTE: if $C' = \emptyset$, then the clause becomes irrelevant and
          doesn't need smoothing, so can be skipped
  \end{itemize}
}

@@

\todo[inline,caption={}]{ Turn these notes into a coherent description (that
  doesn't mention \textsc{Crane}):
  \begin{itemize}
    \item Each base case identified by \FindBaseCases can be interpreted as
          assuming some domain...
    \item \Propagate (\cref{alg2}) to find the transformed formula corresponding
          to each base case identified by \FindBaseCases and call \textsc{Crane}
          on the formula to obtain the required base cases.
    \item Clauses where all literals have variables quantified over the empty
          domain are still removed.
    \item $C'$ is guaranteed to be non-empty.
    \item \Cref{line:smoothing1,line:smoothing2} are explained in
          \cref{sec:smoothingbase}.
  \end{itemize}
}
\todo[inline]{Maybe prove consistency: if two base case overlap, their numerical
  answers also overlap.}

@@

For example, consider the formula
\begin{gather}
  \forall x \in \Delta\text{. }\forall y, z \in \Gamma\text{. }P(x) \lor Q(y, z)\label[formula]{eq:example1}\\
  \forall y, z \in \Gamma'\text{. }Q(y, z)\label[formula]{eq:example2}
\end{gather}
and assume that $\Gamma' \subseteq \Gamma$. If we set $|\Delta|$
to zero and remove clauses with variables quantified over $\Delta$, we get
\begin{equation}\label[formula]{eq:simplified}
  \forall y, z \in \Gamma'\text{. }Q(y, z),
\end{equation}
but the model count of \cref{eq:simplified} is one. However, the actual model
count should be $2^{|\Gamma|^2 - |\Gamma'|^2}$. That is, $Q$ as a relation is a
subset of $\Gamma \times \Gamma$. While \cref{eq:example1} becomes vacuously
true, \cref{eq:example2} fixes the value of $Q$ over
$\Gamma' \times \Gamma' \subseteq \Gamma \times \Gamma$. Hence, the number of
different values that $Q$ can take is
$|(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')| = |\Gamma|^{2} - |\Gamma'|^{2}$.

@@

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Maybe add some complete examples of C++ programs in the supplementary
          material.
          \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by A\&G
            \item re-check submission instructions and formatting guidelines
            \item Full stop at the end of a single-sentence caption?
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
          \end{itemize}
  \end{itemize}
}

>>>>

\crefname{line}{line}{lines}

\crefalias{clause}{equation}
\crefname{clause}{Clause}{Clauses}
\creflabelformat{clause}{#2\textup{(#1)}#3}

\crefalias{formula}{equation}
\crefname{formula}{Formula}{Formulas}
\creflabelformat{formula}{#2\textup{(#1)}#3}

@@

\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator*{\argmin}{argmin} % TODO: soon to be removed

@@

\todo[inline]{Clarify the differences between the input format and the internal
  format}

\paragraph{The \FO{} Logic (After Skolemization and in a special (prenex, CNF)
  form)}
\begin{itemize}
  \item Although existential quantifiers are supported, here we describe the
        format used internally. During preprocessing, all existential
        quantifiers are eliminated using
        Skolemization~\cite{DBLP:conf/kr/BroeckMD14}, and the formula is
        rewritten into a conjunction of clauses, each of which is in
        \emph{prenex normal form}~\cite{hinman2018fundamentals}.
  \item In the spirit of keeping sorts implicit, we always assume formulas `type
        check' with respect to sorts. For example, if $P(x)$, $P(y)$, and
        $x \ne c$ are all part of the formula (for some predicate $P$, variables
        $x$ and $y$, and constant $c$), then $x$, $y$, and $c$ all have the same
        sort.
  \item A \emph{formula} is a conjunction of clauses.
  \item A \emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
        }\forall x_{2} \in \Delta_{2}\dots\text{
        }\forall x_{n} \in \Delta_{n}\text{. }\phi(x_{1}, x_{2}, \dots, x_{n})$,
        where $\phi$ is a disjunction of literals that only contain variables
        $x_{1}, \dots, x_{n}$ (and any constants).
  \item We say that a clause is a \emph{(positive) unit clause} if:
        \begin{itemize}
          \item there is only one literal with a predicate, and
          \item it is a positive literal.
        \end{itemize}
  \item A \emph{literal} is either an atom (i.e., a \emph{positive} literal) or
        its negation (i.e., a \emph{negative} literal).
  \item An \emph{atom} is either:
  \begin{itemize}
    \item $P(t_{1}, \dots, t_{m})$ for some predicate $P/m$ and terms
          $t_{1}, \dots, t_{m}$ or
    \item $x=y$ for some terms $x$ and $y$
  \end{itemize}
  \item An atom is \emph{ground} if it contains no variables (i.e., only
        constants).
  \item The \emph{arity} of a predicate is the number of arguments it takes,
        i.e., $m$ in the case of predicate $P$.
  \item When we want to denote a predicate together with its arity, we write
        $P/m$.
  \item A \emph{term} is either a variable or a constant.
  \item Throughout the paper, we use set-theoretic notation, interpreting a
        formula as a set of clauses and a clause as a set of literals. Moreover,
        for readability, clauses written on separate lines are implicitly
        conjoined.
\end{itemize}

@@

\paragraph{Three Types of Logics}
\begin{itemize}
  \item See \cref{tbl:logics} for a detailed comparison. The notation introduced
        in the table is standard for \Ctwo{}, new for \UFO{}, and redefined to
        be more specific for \FO{}.
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item In many-sorted logic, each term is assigned to a \emph{sort}, and each
        predicate $p/n$ is assigned to a sequence of $n$ sorts. Each sort has
        its corresponding domain. In the input formula, all domains are assumed
        to be pairwise disjoint Most of these assignments are typically left
        implicit and can be reconstructed from the quantifiers. For instance,
        $\forall x,y \in \Delta\text{. }P(x, y)$ implies that variables $x$ and
        $y$ have the same sort. On the other hand, $\forall x \in \Delta\text{.
        }\forall y \in \Gamma\text{. } P(x, y)$ implies that $x$ and $y$ have
        different sorts, and it would be improper to have $x = y$ as part of a
        formula.
  \item
        \begin{itemize}
          \item \FO{} is used as the input format for
                \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
                and its extensions
                \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}~\cite{DBLP:conf/kr/DilkasB23}
                and \Cranetwo{}.
          \item \Ctwo{} is discussed in the literature on
                \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} and related
                methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}
          \item \UFO{} is the input format supported by a version of
                \textsc{FastWFOMC} obtained directly from the authors. Note that
                the publicly available
                version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
                does not support any cardinality constraints.
        \end{itemize}
  \item Note that, in the case of \textsc{ForcLift} and its extensions, support
        for a formula as valid input does not imply that the algorithm will be
        able to compile the formula into a circuit or graph suitable for lifted
        model counting. However, it is known that \textsc{ForcLift} compilation
        is guaranteed to succeed on any \FO{} formula without constants and with
        at most two variables~\cite{DBLP:conf/nips/Broeck11}.
\end{itemize}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item say something about what $\expr{}$ can contain (function calls,
          various arithmetic operations, summations, non-negative integer
          constants, variables introduces by the function signature or
          summations)
  \end{itemize}
}

\begin{itemize}
  \item We write $\expr{}$ for an arbitrary algebraic expression.
  \item Some terms have different meanings in the context of algebra as compared
        to logic. Here, a \emph{constant} is a non-negative integer. Similarly,
        a \emph{variable} is either a parameter of a function or a variable
        introduced by a summation, e.g., $i$ in $\sum_{i=1}^{n} \expr$.
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{\protect\CompileWithBaseCases{$\phi$}}\label{alg:main}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;\label{line:first}
  $\mathcal{E} \gets \Simplify{$\mathcal{E}$}$\;\label{line:second}
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i} \in \mathbb{N}_{0}$}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $\mathcal{E} \gets \mathcal{E} \cup \CompileWithBaseCases{$\psi$}$\;
  }
\end{algorithm}

\Cref{alg:main} outlines our overall approach for compiling a formula into a set
of equation that include the required base cases. In short, we first use the
knowledge compilation algorithm of the original
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} to compile the formula into:
\begin{enumerate*}[label=(\roman*)]
  \item set $\mathcal{E}$ of equations,
  \item map $\mathcal{F}$ from function names to formulas, and
  \item map $\mathcal{D}$ from function names and argument indices to domains.
\end{enumerate*}
After some algebraic simplification, $\mathcal{E}$ is passed to the
\FindBaseCases procedure that returns a set of base cases that we need to find
solutions for (described in \cref{sec:identifying}). For each base case
$f(\mathbf{x})$, we identify the formula associated with $f$ and simplify it
using the \Propagate procedure (described in \cref{sec:simplifying}). The
algorithm then recurses on these simplified formulas and adds the resulting base
case equations to $\mathcal{E}$. \Cref{example:overall} explains \cref{alg:main}
in more detail.

\begin{example}\label{example:overall}
  Let us consider the following formula (previously examined by Dilkas and
  Belle~\shortcite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ to be a
  bijection between two sets $\Gamma$ and $\Delta$:
  \[
    \begin{gathered}
      \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y)\\
      \forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y)\\
      \forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z\\
      \forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z.
    \end{gathered}
  \]
  In particular, we examine the first solution that \Cranebfs{} returns for this
  formula.

  After \cref{line:first,line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain introduced by \Crane. Then
  \FindBaseCases identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both
  cases, \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$
  simplified by assuming that one of the domains is empty. In the first case, we
  recurse on the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$,
  where $S$ is a predicate introduced by Skolemization with weights
  $w^{+}(S) = 1$ and $w^{-}(S) = -1$. Hence, we get the base case
  $g(0, m) = 0^{m}$. In the case of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$}
  returns an empty formula, giving us $g(l, 0) = 1$.
\end{example}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item \Cref{alg2} now supports any base case domain sizes
    \item NOTE: if $C' = \emptyset$, then the clause becomes irrelevant and
          doesn't need smoothing, so can be skipped
  \end{itemize}
}

@@

\todo[inline,caption={}]{ Turn these notes into a coherent description (that
  doesn't mention \textsc{Crane}):
  \begin{itemize}
    \item Each base case identified by \FindBaseCases can be interpreted as
          assuming some domain...
    \item \Propagate (\cref{alg2}) to find the transformed formula corresponding
          to each base case identified by \FindBaseCases and call \textsc{Crane}
          on the formula to obtain the required base cases.
    \item Clauses where all literals have variables quantified over the empty
          domain are still removed.
    \item $C'$ is guaranteed to be non-empty.
    \item \Cref{line:smoothing1,line:smoothing2} are explained in
          \cref{sec:smoothingbase}.
    \item For a literal or a clause $C$, let $\Doms(C)$ denote the set of the
          corresponding domains. Make sure an example illustrates this.
  \end{itemize}
}
\todo[inline]{Maybe prove consistency: if two base case overlap, their numerical
  answers also overlap.}

@@

For example, consider the formula
\begin{gather}
  \forall x \in \Delta\text{. }\forall y, z \in \Gamma\text{. }P(x) \lor Q(y, z)\label[clause]{eq:example1}\\
  \forall y, z \in \Gamma'\text{. }Q(y, z)\label[clause]{eq:example2}
\end{gather}
and assume that $\Gamma' \subseteq \Gamma$. If we set $|\Delta|$
to zero and remove clauses with variables quantified over $\Delta$, we get
\begin{equation}\label[clause]{eq:simplified}
  \forall y, z \in \Gamma'\text{. }Q(y, z),
\end{equation}
but the model count of \cref{eq:simplified} is one. However, the actual model
count should be $2^{|\Gamma|^2 - |\Gamma'|^2}$. That is, $Q$ as a relation is a
subset of $\Gamma \times \Gamma$. While \cref{eq:example1} becomes vacuously
true, \cref{eq:example2} fixes the value of $Q$ over
$\Gamma' \times \Gamma' \subseteq \Gamma \times \Gamma$. Hence, the number of
different values that $Q$ can take is
$|(\Gamma \times \Gamma) \setminus (\Gamma' \times \Gamma')| = |\Gamma|^{2} - |\Gamma'|^{2}$.

@@

\todo[inline]{\Cranebfs{} must be introduced earlier (in the introduction?)}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Maybe add some complete examples of C++ programs in the supplementary
          material.
          \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
            \item Full stop at the end of a single-sentence caption?
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
          \end{itemize}
  \end{itemize}
}
